---
title: "Factors affecting feelings of life"
author: "Ping-ju Kan, Hengguang Zhou, Linhao Liu"
date: "Oct 19, 2020"
output: pdf_document
---

Code and data supporting this analysis is available at: https://github.com/HengguangZhou/sta304A2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(survey)
library(regclass)
gss = read.csv("./gss.csv")

```

## Abstract

We analyzed how different fields give contributions to respondent’s life satisfaction (feelings_life column in data). We applied logistic regression for four fields as input variables and tried to predict whether the respondent felt satisfied or not by checking whether feelings_life greater than or equal to 5. We found that religion and income are two main factors in evaluating life. At the same time, the education level does not contribute a lot. Also, people who live with a partner (no matter married or just living common-law) usually have a more positive feeling of their life.

## Introduction

The feeling of life has a significant influence on your life. A positive feeling of life is related to your physical health and mental health, and it could potentially make you more energetic and more productive in work and study. In this paper, we are interested in how some common factors in life could affect the feeling level of life. With a more advanced understanding of these affecting factors, we could potentially increase our satisfaction level of life.

We focus on four attributes in GSS(General Social Survey) 2017 data set: marital status, income, education and religion participation and how they affect the value of feelings_life, since we assume that these attributes are more or less related to satisfaction of life, based on our daily observation. After our analysis, income, religion participation and marital status verified our assumption, but education does not.

The rest of the paper is organized as follow: In the **Data** section, we filtered needed attributes while keeping all respondents in the dataset. Also, we define feeling_binary as {true: feeling_life >= 5, false: feeling_life < 5}. In the **Model** section, we explained what model we used in detail, and we presented the result in the **Result** section. **Discussion**, **Weakness** and **Next Steps** discussed what we found, what we can improve and what we can do for further investigation.


## Data

GSS(General Social Survey) data is obtained from 20602 respondents, including all non-institutionalized persons 15 years of age and older, living in the 10 provinces of Canada.

We selected five different columns from the GSS 2017 data [5]. It consists of the following: the satisfaction of life for each respondent (feelings_life) as the dependent variable, and the marital status(marital_status), annual income for the respondent(income_respondent), education of the respondent, and the frequency of the participation in religious activities (religious_participation) for the independent variables. 

Population: The population included all persons 15 years of age and older in Canada, excluding: 1. Residents of the Yukon, Northwest Territories, and Nunavut; and 2. Full-time residents of institutions. 

Frame: The survey frame was created using two different components:  Lists of telephone numbers in use (both landline and cellular) available to Statistics Canada from various sources (telephone companies, Census of population, etc.);  The Address Register (AR): List of all dwellings within the ten provinces. Each record in the survey frame was assigned to a stratum within its province. 

Sample: A simple random sample without replacement of records was next performed in each stratum.[6]

The following is the first few lines of the data before the transformation:

```{r}
feeling_life = gss %>% select(feelings_life, marital_status, income_respondent, religion_participation, education)
feeling_life = na.omit(feeling_life)
head(feeling_life)
feeling_binary = feeling_life %>% mutate(feeling_binary = ifelse(feelings_life<5, 0, 1))
for (i in c(1: 6)){  colnames(feeling_binary)[i] = paste(colnames(feeling_binary)[i], "", sep="_") }
```

Since we decided to analyze using a logistic regression model, we first extracted the data we want from the GSS data. Then we transformed the feeling_life from a 0 to 10 numeric categorical variable into a binary variable, depending on whether the value of feelings_life is less than 5 or not, with less than 5 being unsatisfied with life and greater/equal to 5 being satisfied with life.

We think there are some potential drawbacks to this data. For example, we transformed the data for feelings_life from a 0 to 10 discrete numeric categorical variable into a binary variable based on whether the value is less than 5 or not. This might affect the result by amplifying the effect the predictor variables have on the final output since even if the respondents are only borderline satisfied, they would still be considered satisfied with life after the transformation.

Another one would be the fact that we only analyzed the relationship between the satisfaction for life and frequency of religion participation, but there are several more columns from the GSS 2017 data, such as religious importance and whether an individual respondent has an affiliation with one or more religion(s) or not. These drawbacks might potentially affect the results we obtained from performing the logistic regression.


## Model
We want to understand what factor makes the most contribution to being satisfied or unsatisfied in life. In this situation, logistic regression is the best model to use. Logistic regression is suitable when the outcome of interest is binary. One alternative is to use multinomial logistic regression [4]. However, we choose to use binary logistic regression because of the binary nature of our topic.

We make the assumption that the feeling is in linear relationships with the other factors in which we are interested, thus we use a linear model to describe the relationship between the independent variable(feeling_binary) and other factors.

$$
Y_{feeling} = \beta_0+\beta_1X_{education}+\beta_2X_{religion} + \beta_3X_{marital\_status}+\beta_4X_{income}
$$
Here $Y_{feeling}$ represents the binary indicator of satisfaction for life. $X_{education}$ represents
the level of education, $X_{religion}$ represents religion_participation, $X_{marital\_status}$ represents
marital_status, and $X_{income}$ represents the column income_respondent. 
We implement this equation in R using the survey-weighted generalised linear model svyglm() function.

We obtained the population size of Canada for finite population correction[10].

We computed the variance inflation factor to evaluate the linearity between explanatory variables[7]:
```{r}
fpc.srs = rep(36721242, count(feeling_binary))
gss_binary_svy = svydesign(id=~1, data=feeling_binary, fpc=fpc.srs)
logit = svyglm(feeling_binary_ ~ education_ + religion_participation_ + marital_status_ + income_respondent_, gss_binary_svy, family="binomial") 
VIF(logit)
```

We can see that most of the values are low, and thus the explanatory variables are not highly linearly related.

## Results

Package: [9]

```{r, warning=FALSE}
#install.packages("cowplot")
library(cowplot)
# For marital status of respondents who are unsatisfied with life
p1 <- feeling_binary %>%
  filter(feeling_binary_ == 0) %>% 
  group_by(feeling_binary_) %>% 
  count(marital_status_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = marital_status_, group = marital_status_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Marital Status") + ggtitle("Unsatisfied") + coord_polar("y", start=0)
  
```
```{r}
# For marital status of respondents who are satisfied with life
p2 <- feeling_binary %>%
  filter(feeling_binary_==1) %>% 
  group_by(feeling_binary_) %>% 
  count(marital_status_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = marital_status_, group = marital_status_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Marital Status") + ggtitle("Satisfied") + coord_polar("y", start=0)
  
plot_grid(p1, p2)
```
From the resulting pie chart, we can see that the percentage of people that are married and are satisfied with their life is higher than all other marital statuses. In comparison, the percentage of people that are single and unsatisfied with their life are higher than others.

```{r}
# For income of the respondents who are unsatisfied with life
p3 <- feeling_binary %>%
  filter(feeling_binary_==0) %>% 
  group_by(feeling_binary_) %>% 
  count(income_respondent_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = income_respondent_, group = income_respondent_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Income of Respondent") + ggtitle("Unsatisfied") + coord_polar("y", start=0)
  
```

```{r}
# For income of the respondents who are satisfied with life
p4 <- feeling_binary %>%
  filter(feeling_binary_==1) %>% 
  group_by(feeling_binary_) %>% 
  count(income_respondent_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = income_respondent_, group = income_respondent_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Income of Respondent") + ggtitle("Satisfied") + coord_polar("y", start=0)
  
plot_grid(p3, p4)
```
We can see that for the resulting pie chart, almost half the people that are unsatisfied with their life have an income of less than 25000 annually. For the people that are satisfied with their lives, we can see that the people who make less than 25000 are considerably less compared to the percentage of people that are unsatisfied with life. In contrast, the percentage of people who earn between 50000 to 74999 is considerably more.

```{r}
# For education of respondents who are unsatisfied with life
feeling_binary %>%
  filter(feeling_binary_==0) %>% 
  group_by(feeling_binary_) %>% 
  count(education_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = education_, group = education_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Education") + ggtitle("Unsatisfied") + coord_polar("y", start=0)
  
```
For the education of the people who are unsatisfied with their life, the majority of people have an education level of college, high school, or less than a high school diploma.

```{r}
# For education of respondents who are satisfied with life
feeling_binary %>%
  filter(feeling_binary_==1) %>% 
  group_by(feeling_binary_) %>% 
  count(education_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = education_, group = education_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Education") + ggtitle("Satisfied") + coord_polar("y", start=0)
```
We can see that for the people who are satisfied with their lives, the majority of them also have a college, high school, or less than a high school diploma. However, one difference between this chart and the chart for the people that are unsatisfied is that the percentage of people who hold an education degree of either a bachelor or higher than bachelor is considerably higher within the group of people who are satisfied with their life.

```{r}
# For religion participation of respondents who are unsatisfied with life
p7 <- feeling_binary %>%
  filter(feeling_binary_==0) %>% 
  group_by(feeling_binary_) %>% 
  count(religion_participation_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = religion_participation_, group = religion_participation_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Religion participation") + ggtitle("Unsatisfied") + coord_polar("y", start=0)
  
```

```{r}
# For religion participation of respondents who are satisfied with life
p8 <- feeling_binary %>%
  filter(feeling_binary_==1) %>% 
  group_by(feeling_binary_) %>% 
  count(religion_participation_) %>% 
  mutate(percentage=(n/sum(n)) * 100) %>% 
  ggplot(aes(x = feeling_binary_, y = percentage, fill = religion_participation_, group = religion_participation_)) + geom_bar(position="stack", stat="identity") + labs(y = "Percentage", x = "Satisfactory towards life", fill = "Religion participation") + ggtitle("Satisfied") + coord_polar("y", start=0)
  
plot_grid(p7, p8)
```
For the religious participation bar plot, we can see that the percentage of people who do not participate in a religion practice at all is considerably higher in the group of people who are unsatisfied with their life compared to the group of people who are satisfied with their life. Also, the percentage of people who practice religion at least a few times during the year is much higher in the group of people who are satisfied with their lives.

Code source: [8]

```{r}
print(broom::tidy(logit), n = 22)
```
As we can see in the table above, using a 97.5% confidence interval, we found that the religion_participation, marital_status, and income_respondent are closely related to satisfaction. Notably, living common-law, married has a positive impact on happiness, while not participating in religion and having an income less than 25,000 would decrease satisfaction.  On the other hand, we cannot find any significant evidence that supports the relationship between education and satisfaction.

## Discussion

Various research has claimed higher income more or less promotes people’s feelings of happiness[1, 2]. People with higher salaries are able to buy expensive products such as cars, luxuries, new technology goods, and they are capable of enjoying high-quality services. A wealthy person has the ability to satisfy all his or her needs in time. Also, suppose you have a higher salary than someone else. In that case, it usually means you play a more important role in modern society, and it brings superiority complex and satisfaction. Our results are consistent with this conjecture. We found that although the relationship between satisfaction and higher income is not clear, there is strong evidence of a negative relationship between low income and satisfaction.

On the other hand, our results provide strong support that satisfaction is related to religion in a similar way. Religion can also bring a higher level of happiness [3]. Religious activities help the devouts build a broad social network.  The more often you visit the church (religious participation), the more often you meet with priests and friends. 

At the same time, social relationships are also one of the main reasons we believe why having a spouse or partner is important. We can see that being married or living common-law contributes significantly to more life satisfaction in our result. Afterall, spirituality gives meaning to life, enlightenment, inner peace and inspiration. 

In our initial assumption, education is closely related to social status, income and range of knowledge. An educated person usually has a broader range of knowledge, can view things more thoroughly, and has a better understanding of what is going on. So they might be able to find problems in life and solve them more easily. Surprisingly, we found that education level does not have much influence on one’s satisfaction.

One thing worth noting is that only extremes of these factors have a significant impact on satisfaction. By extremes, we mean “having income less than 25000” or “No religion participation at all.” Generally speaking, we do not have significant support for the impact of having middle-range income or a moderate frequency in religion participation to satisfaction. 


## Weaknesses

One weakness of this dataset is that most of the attributes are categorical and thus difficult to analyze.

A possible weakness of our model is using satisfaction level as a binary variable oversimplifies the model and loses information. Simultaneously, we do not have a particularly effective approach for choosing the threshold value while it could be influential to the result.  

## Next Steps

It might be interesting to analyze the correlation between different factors in the future. A surprising result we got is that education is not related to satisfaction. From our perspective, education is likely to have a strong relation to income(or not?), but income is related to satisfaction while education is not. Studying the correlation between them could help us understand their relation to satisfaction better.

Furthermore, there are multiple attributes related to one field in the dataset. For example, income_family and income_respondent are related to income, religion_has_affiliation, religion_importance and religion_participation are all related to religion. We can further investigate these related attributes for more findings.


## References
* 1.Daniel Kahneman and Angus Deaton. High income improves evaluation of life but not emotional well-being. Retrieved August 4, 2010, from https://www.pnas.org/content/pnas/107/38/16489.full.pdf
* 2.Ada Ferrer-i-Carbonell. Income and well-being: an empirical analysis of the comparison income effect. Retrieved from http://darp.lse.ac.uk/papersDB/Ferrer-i-Carbonell_%28JPubE05%29.pdf
* 3.: Religion’s Relationship to Happiness, Civic Engagement and Health Around the World. Retreved from https://www.pewforum.org/2019/01/31/religions-relationship-to-happiness-civic-engagement-and-health-around-the-world/
* 4: STA304 - Logistic Regression Intro, Samantha-Jo Caetano. Retrieved from week 4 course material.
* 5: Statistics Canada. (2020, April). GSS 2017 Dataset, Retrieved from http://dc.chass.utoronto.ca/myaccess.html
* 6: Statistics Canada. (2020, April). GSS 2017: User Guide for the Public Use Microdata File (PUMF). Retrieved from https://sda-artsci-utoronto-ca.myaccess.library.utoronto.ca/sdaweb/dli2/gss/gss31/gss31/more_doc/GSS31_User_Guide.pdf
* 7: Government of Canada, Statistics Canada. (2020, September 29). Population estimates, quarterly. Retrieved October 17, 2020, from https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710000901
* 8: Table format code from https://piazza.com/class/kevvm6h64jo5hh?cid=255
* 9: Cowplot package in R. Retrieved from  https://cran.r-project.org/web/packages/cowplot/index.html
* 10: regclass package in R. Retrieved from https://www.rdocumentation.org/packages/regclass/versions/1.6/topics/VIF






